---
title: "R's 3 Plotting Systems"
output: html_notebook
---


There are three major plotting systems in R:
1 Base graphics
2 Lattice graphics
3 ggplot2 

Additionally, there is a significant amount of development work going on to allow R users to produce interactive Javascript based plots. In this module we ill give an example of plotting histograms in all three systems.

### The Data
For this we will use Duncan's famous `Prestige` data set in the `car` package that shows income, education level and a prestige score for a number of workers classified as being either "professional" (prof), "blue collar" (bc), or "white collar" (wc). We will sho how to plot histograms of income for all three categories laid out side-by-side for comparison

### Base Graphics Histograms

In this first block of code we load the data set which is organized as a data frame and break out the data for each category into separate vectors. We do this because the base R hist function expects a vector as input.
```{r}
library(car)
data(Duncan)
dim(Duncan)
head(Duncan)
# Select data using basic subsetting
inc_prof <- Duncan[Duncan$type=="prof","income"]
inc_bc <- Duncan[Duncan$type=="bc","income"]
inc_wc <- Duncan[Duncan$type=="wc","income"]
```  

In this next block of code we first use the `par` command to set up a "panel" on the graphics output device to contain three side-by-side plots, then we plot each histogram one by one.


```{r}
par(mfrow=c(1,3))  # set option to put 3 plots in one pane
hist(inc_bc, prob = TRUE, col = "pink", main = ("Income BC"), xlab=("Dollars"))
             lines(density(inc_bc))
hist(inc_prof, prob = TRUE, col = "yellow", main = ("Income Prof"), xlab=("Dollars")) 
             lines(density(inc_prof))
hist(inc_wc, prob = TRUE, col = "light blue", main = ("Income WC"), xlab=("Dollars")) 
lines(density(inc_wc))
par(mfrow=c(1,1))
```   



### Lattice (trellis) Graphics
Lattice graphics are the second major plotting system in R. Plots built with lattice have a very distinctive look, but the real value is the ease of making trellis plots - graphs that display a variable conditioned on an other variable. Some useful websites are: 
http://www.statmethods.net/advgraphs/trellis.html http://user2007.org/program/presentations/sarkar.pdf

You can see that lattice has a `panel` parameter that can be set in the histogram. Also notice tht the `histogram()` function expects a data frame as input.
```{r}
library(lattice)
histogram( ~ income | type, 
           data = Duncan,
           nint=10,
           xlab = "Income",  
           main= "Hitogram by profession",
           type = "density",
           panel = function(x, ...) {
             panel.histogram(x, ...)
             panel.mathdensity(dmath = dnorm, col = "black",
                               args = list(mean=mean(x),sd=sd(x)))
           } )
```     

### ggplot2 Graphics

ggplot is the third major plotting system for R. It is based on Leland Wilkinson’s grammar of graphics. Plots are built in layers.
Some useful websites are: 
http://www.cs.uic.edu/~wilkinson/TheGrammarOfGraphics/GOG.html 
https://ggplot2.tidyverse.org/
http://www.cookbook-r.com/Graphs/
https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf

```{r}
library(ggplot2)
p1 <- ggplot(Duncan,aes(income,..density.., fill=type))
p1 + geom_histogram(bins=10) + 
     facet_grid(. ~ type) +
     geom_density(alpha = .5) +
     xlab("Income: Canadian $") +
     ggtitle("Histogram by Profession")
```   



